\LARGE \bf Enriching Robot's Actions with Affective Movements

### 2017-03-08 RO-MAN-2017-2/Parts/EmotionalTree.tex
new changes

Emotional Execution Tree is a connected acyclic graph $G(V,E)$ with $|V|$ vertexes and $|E|$ edges. The root and non-leaf nodes could be of either \textit{parallel} or \textit{sequential} type. The parallel node could be one out of four different sub-types: (i) action and emotion synchronous, (ii) action synchronous and emotion asynchronous, (iii) action asynchronous and emotion synchronous, or (iv) action and emotion asynchronous. Sequential nodes could just be one of two sub-types: emotion synchronous or asynchronous. Action synchronous means that each time that a parallel node receives a ``finish'' notification (i.e., success or failure), it will broadcast the message to all nodes that derived it and to its predecessor. If a sequence node receives a finish message, it will execute the next branch. When all branches have been executed, it communicates the end of the action. On the other hand, emotion synchronous means that each time that a node (either sequence or parallel) receives an emotion synchronization message, it will propagate the message to all branches to move to the consecutive emotional expression. If a node is principal and it has finished to execute all actions, it will notify its predecessor.

This distinction creates the possibility to synchronize emotional changes without affecting the normal execution of an action and it also enables synchronization among parallel actions. Finally, leaf nodes could only be simple action nodes that have been implemented in the system. Any node can belong to one of two levels: principal or  secondary. If a node is principal, it will notify its predecessor about the messages that it has received, while the secondary node cannot propagate any message to its predecessor.%Why is this needed?
Compound actions are implemented combining all type of nodes.

>>>>

Emotional Execution Tree is a connected acyclic graph $G(V,E)$ with $|V|$ vertexes and $|E|$ edges. The root and non-leaf nodes could be of either \textit{parallel} or \textit{sequential} type. Sequential nodes execute one branch after the other. Once all the branches have been executed, it notifies to its predecessor to continue with the execution of other nodes.  The parallel node offers two type of synchronization. one is at action and the other at emotion. This distinction between action and emotion synchronization enables the possibility to change emotion parameter without affecting the action. done could be one out of four different sub-types: (i) action and emotion synchronous, (ii) action synchronous and emotion asynchronous, (iii) action asynchronous and emotion synchronous, or (iv) action and emotion asynchronous. 
 Action synchronous means that each time that a parallel node receives a ``finish'' notification (i.e., success or failure), it will broadcast the message to all nodes that derived it and to its predecessor, which will stop their execution.

 If a sequence node receives a finish message, it will execute the next branch. When all branches have been executed, it communicates the end of the action. On the other hand, emotion synchronous means that each time that a node (either sequence or parallel) receives an emotion synchronization message, it will propagate the message to all branches to move to the consecutive emotional expression. If a node is principal and it has finished to execute all actions, it will notify its predecessor.

This distinction creates the possibility to synchronize emotional changes without affecting the normal execution of an action and it also enables synchronization among parallel actions. Finally, leaf nodes could only be simple action nodes that have been implemented in the system. Any node can belong to one of two levels: principal or  secondary. If a node is principal, it will notify its predecessor about the messages that it has received, while the secondary node cannot propagate any message to its predecessor. This enables the possibility to have actions that are the ones to command over the others. 
Compound actions are implemented combining all type of nodes.

### 2017-03-08 RO-MAN-2017-2/Parts/Introduction.tex
new changes

In theory, the projection of emotion with humanoid embodiments could be simplified to mimic the same movements that humans. However, this idea is not possible due robots physical limitations. Therefore, an exact matching of humans’ movements to convey emotions could not be used in robots \cite{Saerbeck2007,Canamero2010}. Therefore diverse researchers are studying diverse features and values to express emotions with different platforms. As a consequence, these results could not be widely used due to the differences of the platforms.

This paper presents an Emotional Enrichment System (EES), which modify actions' parameters and add additional actions to create the illusion of emotion expression in a robot. Although the EES was originally conceived to be used in an autonomous performance robot~\cite{angel2013} to enrich actions with emotions, its design was devised to make it extendable to other platforms and adaptable to new tasks. To achieve this goal, the system relies on an Emotional Execution Tree (EXT), which is based on simple actions, sequential and parallel nodes. Additionally, it is used the concept of compound actions to group a bunch of nodes, which reduces the tree dimension and allows the reuse of recurrent actions  generated by specific combinations of simple actions and other nodes. This EXT has been formalized to give a guideline to further implementations and extensions.

>>>>

In theory, the projection of emotion with humanoid embodiments could be simplified to mimic the same movements that humans. However, this idea is not possible due robots physical limitations. Therefore, an exact matching of humans' movements to convey emotions could not be used in robots \cite{Saerbeck2007,Canamero2010}. Therefore diverse researchers are studying diverse features and values to express emotions with different platforms. As a consequence, these results could not be widely used due to the differences of the platforms.

This paper presents an Emotional Enrichment System (EES), which modify actions' parameters and add additional actions to create the illusion of emotion expression in a robot. Although the EES was originally conceived to be used to facilitate the study of emotion projection, its design was devised to make it extendable to other platforms and adaptable to new tasks. To achieve this goal, the system relies on an Emotional Execution Tree (EXT), which is based on simple actions, sequential and parallel nodes. Additionally, it is used the concept of compound actions to group a bunch of nodes, which reduces the tree dimension and allows the reuse of recurrent actions  generated by specific combinations of simple actions and other nodes. This EXT has been formalized to give a guideline to further implementations and extensions.

### 2017-03-08 RO-MAN-2017-2/Parts/RelatedWork.tex
new changes

The use of emotion enrichment to improve human robot interaction is not new trend. There have been several works that have enhance their social robots with emotions or study how to convey them in robotic platforms~\cite{Li2011,Brown2014}. One of the most well-known expressive robots is Kismet~\cite{Breazeal2002}, a robotic face able to interact with people and show emotions. This platform uses a specific set of movements based on the Ekman's studies on human emotion expression~\cite{Ekman2004}. Other approaches have tried to use anthropomorphic~\cite{Arras2012} and human-like platforms to convey emotions to study the response of people towards the robot. However, the emotions portrayed were hand-coded, hard-wired to the respective platforms, and their parametrization was not available.\\
On the other hand, studies focused on entertainment robotics have tried to introduce emotional actions to improve the audience's experience. Breazeal and collaborators~\cite{Breazeal2003} used one robot on the stage. This anemone-like robot had few behaviours, which included getting scared when a person comes too close; the robot was able to show some basic emotions (i.e., fear and interest). Knight~\cite{Knight2011b,Knight2010} used the platform NAO to produce a sort of stand-up comedy. The robot performs basic actions to add some expressiveness to the joke, but it is not intended to project any emotion. Trying to add some theatrical realism, Breazeal and collaborators~\cite{Breazeal2008} designed and implemented a system to control a lamp. The main characteristic of this lamp is that it could be controlled by just one person, which selected pre-coded emotions.\\
Other works in performance robotics have developed systems that do not convey any kind of emotion as \textit{Roboscopie}~\cite{Roboscopie2012,Lemaignan2012}, Fan and collaborators~\cite{Fan2009,Fan2013}, and adaptation of Shakespeare's  Midsummer Night's Dream~\cite{murphy2011} uses robots in performances with real actors but without any emotion expression.\\
Although these works use emotions, their main focus was the interpretation of postures or just the use of emotions to increase their robot appealing, but none of them considered the importance of emotional system that could be used by others. As a consequence, most of these works have created emotional systems that could just work in their specific system.

>>>>

The use of emotion enrichment to improve human robot interaction is not new trend. There have been several works that have enhance their social robots with emotions or study how to convey them in robotic platforms~\cite{Li2011,Brown2014}. One of the most well-known expressive robots is Kismet~\cite{Breazeal2002}, a robotic face able to interact with people and show emotions. This platform uses a specific set of movements based on the Ekman's studies on human emotion expression~\cite{Ekman2004}. Other approaches have tried to use anthropomorphic~\cite{Arras2012} and human-like platforms to convey emotions to study the response of people towards the robot. However, the emotions portrayed were hand-coded, hard-wired to the respective platforms, and their parametrization was not available.

On the other hand, studies focused on entertainment robotics have tried to introduce emotional actions to improve the audience's experience. Breazeal and collaborators~\cite{Breazeal2003} used one robot on the stage. This anemone-like robot had few behaviours, which included getting scared when a person comes too close; the robot was able to show some basic emotions (i.e., fear and interest). Knight~\cite{Knight2011b,Knight2010} used the platform NAO to produce a sort of stand-up comedy. The robot performs basic actions to add some expressiveness to the joke, but it is not intended to project any emotion. Trying to add some theatrical realism, Breazeal and collaborators~\cite{Breazeal2008} designed and implemented a system to control a lamp. The main characteristic of this lamp is that it could be controlled by just one person, which selected pre-coded emotions.

Other works in performance robotics have developed systems that do not convey any kind of emotion as \textit{Roboscopie}~\cite{Roboscopie2012,Lemaignan2012}, Fan and collaborators~\cite{Fan2009,Fan2013}, and adaptation of Shakespeare's  Midsummer Night's Dream~\cite{murphy2011} uses robots in performances with real actors but without any emotion expression.\\
Although these works use emotions, their main focus was the interpretation of postures or just the use of emotions to increase their robot appealing, but none of them considered the importance of emotional system that could be used by others. As a consequence, most of these works have created emotional systems that could just work in their specific system.

### 2017-03-08 RO-MAN-2017-2/Parts/description.tex
new changes

	\item \textit{Emotion description} gives information of the parameters that should be changed in all the simple actions to express the desired emotion. Therefore, for each combination emotion-action should be given a description of how the parameters are defined. If there is any action that does not have any specification, the system will not change its parameters. The following is the EBFN generated to modify movement parameters: 
\begin{grammar}
[(colon){$\models$}]
[(semicolon)$|$]
[(comma){}]
[(period){\\}]
[(quote){\begin{bf}}{\end{bf}}]
[(nonterminal){$\langle$}{$\rangle$}]
<emotion description>: "'\{'","'emotion:'",<string> "','","'observation:'",<string> "','",<action>"'(,'",<action>)* "'\}'".
<action>:<string>"':\{'",<description> "','",<actions affected> "'\}'".
<description>:"'description:\{'",<action name>"','",<emotion>"','"\\
<parameter's type>"'\}'".
<action name>:"'emotionProfileAction:'",<string>.
<emotion>:"'emotionProfileEmotion:'",<string>.
<parameter's type>: "'movement\_parameter'".
<actions affected>:"'actions:\{'",<action parameter>,\\("','"<action parameter>)*"'\}'".
<action parameter>: <string>"':\{'",<reference>"','"\\<repetition>"','",<parameters>"'\}'".
<reference>: "'reference:'",<number>.
<repetition>: "'repetition:'","'yes'";"'no'".
<parameters>: "'parameters:['" <parameter description>\\("','",<parameter description>)* "']'".
<parameter description>: <movement parameter description>;<new parameters>.
<movement parameter description>:"'\{time:'",<number>"',space:'"\\<number>"'\}'".
\end{grammar}

	\item \textit{Character's emotions} parameters give the system information about character's ''rhythm'' for each pair action-emotion. In other words, it is going to say how the emotional parameters for the desired action should be modulated. If there is not any information, the system will use the default emotional values for the action. The following is the corresponding EBFN:
\begin{grammar}
[(colon){$\models$}]
[(semicolon)$|$]
[(comma){}]
[(period){\\}]
[(quote){\begin{bf}}{\end{bf}}]
[(nonterminal){$\langle$}{$\rangle$}]
<character emotion profile>:"'\{'" <emotion>"','",<actions>"'\}'".
<emotion>: "'emotion:'",<string>.
<actions>: "'actions:['", <action>("','",<actions descriptor>)* "']'".
<action>: "'\{ action: '", <string> "','","'bias:'", <number> "','"\\" 'amplitude:'", <number> ", 'long:'",<number> "\}". 
\end{grammar}

	\item \textit{Action message} contains all the necessary information for to execute an action. The following is a script's EBNF:
\begin{grammar}
[(colon){$\models$}]
[(semicolon)$|$]
[(comma){}]
[(period){\\}]
[(quote){\begin{bf}}{\end{bf}}]
[(nonterminal){$\langle$}{$\rangle$}]
<action>: <simple action>;<compound action>;<context>.
<simple action>: "'\{'" <action header> "','","'parameters:['" <simple action parameters> "']\}'".
<compound action>: "'\{'" <action header> "','","'parameters:['" <compound action parameters> "']\}'".
<action header>: "'type:'", <action type> "','","'name: '" <string>  "','", <is primary>.
<simple action parameters>: "'\{'"<parameter header> "','" ,<parameter description> "'\}'". 
<compound action parameters>:  <simple action parameters> ("','", <simple action parameters>)* .
<context>:"'\{'" <action type> "','", <emotion sync> "','", <action sync> "','", <is primary> "','"\\" 'information:'", <string> "','"," 'actions: '", <action> "'\}'".
<parameter header>: "'type:'", <parameter type> "','"," 'name:'",<string>.
<parameter description>: <parameter amplitude>;<parameter circle>;<parameter landmark>;\\<parameter point>;<parameter speech>;\\<parameter square>;<parameter time>.
<parameter type>: "'mandatory\_parameter'";"'optional\_parameter'". 
<is primary>: "'yes'";"'no'".
<emotion sync>: "'yes'";"'no'".
<action type>: "'parallel\_context'";"'serial\_context'";"'simple\_action'";\\"'composite\_action'".
\end{grammar}
\end{itemize}

>>>>

	\item \textit{Emotion description} gives information of the parameters that should be changed in all the simple actions to express the desired emotion. Therefore, for each combination emotion-action should be given a description of how the parameters are defined and how they change in time. 
%If there is any action that does not have any specification, the system will not change its parameters. 
The first version of this parameters is done for movement. Therefore, it was considered four descriptors for this type of parameter. The first is the reference, which inform the system the velocity used to calculate the parameter. This reference is used to let the modulation of the parameters depending on the character. The second descriptor is the tuple space and time. with this tuple is possible to calculate the velocity in which the movement must be done in order to convey the desire emotion. in order to enable the possibility to have different velocities during the execution, it could be described a sequence of this tuple. in this case, the space descriptor plays an important role, which is to inform the length in which the desired velocity must be used. This bring the last description which is the repetition that is used if there is more than one tuple and it should be repeat during the duration of the movement. The EBFN to described these descriptors in the system is the following: 
\begin{grammar}
[(colon){$\models$}]
[(semicolon)$|$]
[(comma){}]
[(period){\\}]
[(quote){\begin{bf}}{\end{bf}}]
[(nonterminal){$\langle$}{$\rangle$}]
<emotion description>: "'\{'","'emotion:'",<string> "','","'observation:'",<string> "','",<action>"'(,'",<action>)* "'\}'".
<action>:<string>"':\{'",<description> "','",<actions affected> "'\}'".
<description>:"'description:\{'",<action name>"','",<emotion>"','"\\
<parameter's type>"'\}'".
<action name>:"'emotionProfileAction:'",<string>.
<emotion>:"'emotionProfileEmotion:'",<string>.
<parameter's type>: "'movement\_parameter'".
<actions affected>:"'actions:\{'",<action parameter>,\\("','"<action parameter>)*"'\}'".
<action parameter>: <string>"':\{'",<reference>"','"\\<repetition>"','",<parameters>"'\}'".
<reference>: "'reference:'",<number>.
<repetition>: "'repetition:'","'yes'";"'no'".
<parameters>: "'parameters:['" <parameter description>\\("','",<parameter description>)* "']'".
<parameter description>: <movement parameter description>;<new parameters>.
<movement parameter description>:"'\{time:'",<number>"',space:'"\\<number>"'\}'".
\end{grammar}

	\item \textit{Character's emotions} parameters give the system information about character's ''rhythm'' for each pair action-emotion. To describe this changes, it was decided to used bias, amplitude and long as descriptors. Bias refers to a constant value that is added to all the sequence of emotional parameters. Amplitude amplify the velocity respect the bias if the number is bigger than one, or attenuate if it is between zero and one. 
%TODO if there is enough space add an imagine explaining these concepts

	\item \textit{Action message} contains all the necessary information for to execute an action. The following is the EBNF of an action:
\begin{grammar}
[(colon){$\models$}]
[(semicolon)$|$]
[(comma){}]
[(period){\\}]
[(quote){\begin{bf}}{\end{bf}}]
[(nonterminal){$\langle$}{$\rangle$}]
<action>: <simple action>;<compound action>;<context>.
<simple action>: "'\{'" <action header> "','","'parameters:['" <simple action parameters> "']\}'".
<compound action>: "'\{'" <action header> "','","'parameters:['" <compound action parameters> "']\}'".
<action header>: "'type:'", <action type> "','","'name: '" <string>  "','", <is primary>.
<simple action parameters>: "'\{'"<parameter header> "','" ,<parameter description> "'\}'". 
<compound action parameters>:  <simple action parameters> ("','", <simple action parameters>)* .
<context>:"'\{'" <action type> "','", <emotion sync> "','", <action sync> "','", <is primary> "','"\\" 'information:'", <string> "','"," 'actions: '", <action> "'\}'".
<parameter header>: "'type:'", <parameter type> "','"," 'name:'",<string>.
<parameter description>: <parameter amplitude>;<parameter circle>;<parameter landmark>;\\<parameter point>;<parameter speech>;\\<parameter square>;<parameter time>.
<parameter type>: "'mandatory\_parameter'";"'optional\_parameter'". 
<is primary>: "'yes'";"'no'".
<emotion sync>: "'yes'";"'no'".
<action type>: "'parallel\_context'";"'serial\_context'";"'simple\_action'";\\"'composite\_action'".
\end{grammar}
\end{itemize}

### 2017-03-08 RO-MAN-2017-2/RO-MAN2017-2.tex
new changes

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Emotions are considered by many researches as a characteristic that could be beneficial in social robotics, since they enrich human-robot interaction with non-verbal clues. Although there have been works that have studied emotion expression in robots, mechanism to generate emotion are highly integrated with the rest of the system. This unable the possibility to use their approaches in different applications. This paper present a system that has been initially created for a theatrical robot to enrich with emotions its actions, but it has been designed to enable the possibility to be adapted to be used in other fields. The emotional enrichment system has been envisioned to be used with a action decision system. A formal description of the system is provided to be used as a formal reference for further extensions and possible modifications. The system has been adapted to two different platforms with different grades of freedom: Keepon and Triskarino. 
\end{abstract}

\section{Introduction}
\input{./Parts/Introduction}
\section{Related Work}
\input{./Parts/RelatedWork}
\section{Basic Concepts}
\label{sec:concepts}
The system is based on six main concepts: \textit{simple actions}, \textit{compound actions}, \textit{action message}, \textit{emotional descriptors}, \textit{character description} and \textit{emotional execution tree}. \textit{Action message} establishes the structure of the message to describe any kind of action (i.e., simple and compound). This message also specifies how the actions are executed (i.e., in parallel or in sequence) and which action is predominant (i.e., primary or secondary). \textit{Emotional parameters} describe how the emotional enrichment should be done to convey a specific emotion in a specific simple action. This description could also include addition of other simple actions. \textit{Character description} enables the possibility to establish how to modify emotional expressions to generate diverse treats. Finally, \textit{Emotional Execution Tree} is a computational representation of desired actions that should be executed. This tree is first created from the action message description and then modified using the emotional parameters and character description.  
\subsection{Simple and Compound Actions}
\input{./Parts/SimpleActions} 
\section{Describing Components}
\input{./Parts/description}
\section{Emotional Tree}
\input{./Parts/EmotionalTree}
\section{Implementation}
\input{./Parts/Implementation}
\section{Conclusions and Further work}
An Emotional Enrichment System has been designed and implemented to enrich robots' movements with emotions. To achieve this, it was used an Emotional Execution Tree created from three different types of nodes: simple actions, parallel, and sequential. Simple actions are functions that map a set of parameters to specific movements. Sequential executes in order the sequence of actions associated to this type of node, while parallel executes them all at the same time. To enable synchronization among simple actions, parallel nodes could be one of four different subtypes, and sequential just one of two different subtypes. A formalization of the Emotional Execution Tree and the principal consideration during the implementation of the system have also been provided. To show the system's versatility, it was used with quite different platforms such as a Keepon Pro and Triskarino. Keepon was used to perform simple actions, while Triskarino was used to test complex actions with different parametrizations of emotions.\\
The results obtained show that the system could enrich the robotic actions with emotions, which could be parametrized from configuration files. Additionally, the design of the system makes it possible to  adopt it for different platforms using the same action description.
%\input{./Parts/Conclusions}
\bibliographystyle{IEEEtran}
\bibliography{Bibliography,BibloNew,Biblography}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Emotions are considered by many researches as a characteristic that could be beneficial in social robotics, since they enrich human-robot interaction with non-verbal clues. Although there have been works that have studied emotion expression in robots, mechanism to generate emotion are highly integrated with the rest of the system. This unable the possibility to use their approaches in other applications. This paper present a system that has been initially created to facilitate the study of emotion projection, but it was designed to enable its adaptations in other fields. The emotional enrichment system has been envisioned to be used with any action decision system. A description of the system components and its characteristics are provided. The system has been adapted to two different platforms with different grades of freedom: Keepon and Triskarino. 
\end{abstract}

\section{Introduction}
\input{./Parts/Introduction}
\section{Related Work}
\input{./Parts/RelatedWork}
\section{Basic Concepts}
\label{sec:concepts}
The system is based on six main concepts: \textit{simple actions}, \textit{compound actions}, \textit{action message}, \textit{emotional descriptors}, \textit{character description} and \textit{emotional execution tree}. \textit{Action message} establishes the structure of the message to describe any kind of action (i.e., simple and compound). This message also specifies how the actions are executed (i.e., in parallel or in sequence) and which action is predominant (i.e., primary or secondary). \textit{Emotional parameters} describe how the emotional enrichment should be done to convey a specific emotion in a specific simple action. This description could also include addition of other simple actions and vary over time. \textit{Character description} enables the possibility to establish how to modify emotional expressions to generate diverse treats. Finally, \textit{Emotional Execution Tree} is a computational representation of desired actions that should be executed. This tree is first created from the action message description and then modified using the emotional parameters and character description.  
\subsection{Simple and Compound Actions}
\input{./Parts/SimpleActions} 
\section{Describing Components}
\input{./Parts/description}
\section{Emotional Tree}
\input{./Parts/EmotionalTree}
\section{Implementation}
\input{./Parts/Implementation}
\section{Conclusions and Further work}
An Emotional Enrichment System has been designed and implemented to enrich robots' movements with emotions. To achieve this, it was used an Emotional Execution Tree created from three different types of nodes: simple actions, parallel, and sequential. Simple actions are functions that map a set of parameters to specific movements. Sequential executes in order the sequence of actions associated to this type of node, while parallel executes them all at the same time. To enable synchronization among simple actions, parallel nodes could be one of four different subtypes, and sequential just one of two different subtypes. A formalization of the Emotional Execution Tree and the principal consideration during the implementation of the system have also been provided. To show the system's versatility, it was used with quite different platforms such as a Keepon Pro and Triskarino. Keepon was used to perform simple actions, while Triskarino was used to test complex actions with different parametrizations of emotions.\\
The results obtained show that the system could enrich the robotic actions with emotions, which could be parametrized from configuration files. Additionally, the design of the system makes it possible to  adopt it for different platforms using the same action description.
%\input{./Parts/Conclusions}
\bibliographystyle{IEEEtran}
\bibliography{Bibliography,BibloNew,Biblography}
